---
title: "Multilevel models and more on computation"
subtitle: "Lab 6, CSSS 564"
author: "Connor Gilroy"
output: 
  html_document:
    toc: true
    toc_float: true
    fig_width: 5
    fig_height: 3
---

# Goal

- Understand the concept and uses of multilevel models
- Fit multilevel models in Stan and understand the syntax
- Address computational challenges of MCMC by adjusting the models

# Relevant readings

Why use multilevel models?

http://elevanth.org/blog/2017/08/24/multilevel-regression-as-default/

What's the algebra?

http://elevanth.org/blog/2017/09/07/metamorphosis-multilevel-model/

What diagnostics are there for making sure your model fit alright?

http://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html

See also: 

- Kruschke, Chapter 9 (he calls these "hierarchical models")
- McElreath, Chapter 12 (pdf here: http://xcelab.net/rmpubs/rethinking/Statistical_Rethinking_sample.pdf)

# Introduction

Multilevel models are about the *partial pooling* of information. They're useful whenever data are organized into clusters or groups. For instance:

- multiple observations per individual (in medicine, psychology, sports, panel data...)
- classrooms, schools, districts in education data
- geographic regions
- related demographic categories, especially in political or opinion polling---multilevel regression and poststratification (MRP or "Mister P") is quite common
- studies or polls in meta-analyses

# Setup

```{r message=FALSE, warning=FALSE}
library("rstan")
library("tidyverse")
library("bayesplot")

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
theme_set(theme_minimal())
knitr::opts_chunk$set(fig.align = "center")
```

# Model 1: Binary outcome, no covariates

We start with a simple binary outcome model with a link function: 

$$y_i \sim \text{Bernoulli}(p_i)$$
$$p_i = \text{logit}^{-1}(\alpha_{unit[i]})$$

What makes it multilevel is that those $\alpha$ values come from a distribution: 

$$\alpha_j \sim \text{Normal}(a, \tau)$$

We can put some fairly generic weakly informative priors on the parameters of that distribution: 

$$a \sim \text{Normal}(0, 10)$$
$$\tau \sim \text{Cauchy}^+(0, 1)$$

## The data: contraception in Bangladesh

This model uses data from Richard McElreath's `rethinking` package, and implements the model from the second blog post above. Install the package if you need to.

```{r}
# devtools::install_github("rmcelreath/rethinking")
data("bangladesh", package = "rethinking")

# subset the data
set.seed(123)
bangladesh <- 
  as_tibble(bangladesh) %>% 
  sample_n(100)

# look at the data
glimpse(bangladesh)
```

The data are on women's contraceptive use in Bangladesh. The outcome `use.contraception` is binary, and each woman lives in a `district`. We won't use the other variables.

We're subsetting the data because the computational problem isn't so obvious with the full data.

We created a problem when we sampled, though! Since we sampled, we no longer know if our districts are represented by integers from 1 to J.

```{r}
unique(bangladesh$district)
```

How can we fix that? One way is by playing with data types: 

```{r}
unique(as.integer(as.factor(bangladesh$district)))
```

With that in mind, let's format the data for Stan:

```{r}
d <- list(
  y = bangladesh$use.contraception,
  district = as.integer(as.factor(bangladesh$district))
)

# Indices
d$N <- length(d$y)
d$J <- length(unique(d$district))
```


## Multilevel model syntax

Let's have a closer look at what's going on with all those indices: 

```{stan eval=FALSE, output.var=""}
data {
  ...
  // number of districts
  int<lower = 1> J;
  // district index
  int<lower = 1, upper = J> district[N];
  ...
}
...
transformed parameters {
  vector<lower = 0, upper = 1>[N] p;
  for (i in 1:N) {
    p[i] = inv_logit(alpha[district[i]]);
  }
}
...
```

Note: this *isn't* the most efficient way to fit a logit model in Stan! There's a shortcut function `bernoulli_logit`: 

```{stan eval=FALSE, output.var=""}
model {
  ...
  // likelihood
  for (i in 1:N) {
    y[i] ~ bernoulli_logit(alpha[district[i]]);
  }
}
```

## The centered version of the model

```{r message=FALSE}
fit_cp <- stan("stan/multilevel_logit_cp.stan", data = d, chains = 2)
```

```{r}
print(fit_cp, pars = "tau")
```

```{r}
mcmc_trace(as.array(fit_cp), pars = "tau")
```

## The noncentered version

Rewrite the link part like this: 

$$\text{logit}(p_i) = a + \tau * z_{unit[i]}$$

Then the distribution for the random effects becomes this:

$$z_j \sim \text{Normal}(0, 1)$$

This can be easier to sample from using HMC. In Stan, we'd change the model to this: 

```{stan eval=FALSE, output.var=""}
...
parameters {
  vector[J] z;
  real a;
  real<lower = 0> tau;
}
transformed parameters {
  vector<lower = 0, upper = 1>[N] p;
  for (i in 1:N) {
    p[i] = inv_logit(a + tau * z[district[i]]);
  }
}
model {
  // priors
  z ~ normal(0, 1);
  ...
}
```

```{r message=FALSE}
fit_ncp <- stan("stan/multilevel_logit_ncp.stan", data = d, chains = 2)
```

```{r}
print(fit_ncp, pars = "tau")
```

```{r}
mcmc_trace(as.array(fit_ncp), pars = "tau")
```

# Model 2: Continuous outcome, with covariates

## Data

`nlschools` is a data set of test scores of 8th-grade pupils from different schools in the Netherlands. (`?MASS::nlschools` for more information)

```{r}
data("nlschools", package = "MASS")

# sample a subset
set.seed(123)
nlschools <- 
  as_tibble(nlschools) %>% 
  sample_n(200)

# look at the data
glimpse(nlschools)
```

We'll model language test scores `lang` as a function of class size `GS` and family socio-economic status `SES`.

```{r}
# format the data into a list for Stan
d_nls <- list()

# vector of scaled outcome
d_nls$y <- as.vector(scale(nlschools$lang))

# matrix of scaled covariates
d_nls$X <- 
  nlschools %>%
  select(GS, SES) %>%
  scale() %>%
  as.matrix()

# numeric index from 1..J for the groups
d_nls$group <- as.integer(as.factor(as.integer(nlschools$class)))
```

All the indices! 

```{r}
d_nls$N <- length(d_nls$y)
d_nls$K <- ncol(d_nls$X)
d_nls$J <- length(unique(d_nls$group))
```

## Fit the model

Have a look at `multilevel_regression_cp.stan` and `multilevel_regression_ncp.stan` in the `stan` folder.

First, fit the centered version: 

```{r message=FALSE}
# fit the centered version of the regression
```

```{r}
# check model summaries and trace plots for model coefficients 
```

Then, fit the noncentered version:

```{r message=FALSE}
# fit the noncentered version of the regression
```

```{r}
# check model summaries and trace plots for model coefficients 
```
